---
title: "Post-Harmonization Downstream Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-Harmonization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

After harmonization, multiple-sites batch effect get controlled (or ideally eliminated). Some studies require further investigation of life span age trend of brain structures or other significant variable effects on brain structures. ComBatFamQC provides a post harmonization tool to:

-   interactively visualize age trend: `age_shiny`
-   generate residuals eliminating specific covariates' effects: `residual_gen`
    -   generate residuals from scratch
    -   generate residuals based on existing regression model

# Set up

Import `ComBatFamQC` package and read in harmonized data set for age trend visualization. We use `age_df` data for age trend visualization in the vignette. To be noticed the read-in data set should be a data frame (not tibble). 
 
```{r setup, eval = FALSE}
library(ComBatFamQC)
data(age_df)
```

# Life Span Age Trend Visualization

In this step, we need to generate a list of data sets for all rois. Each roi's data set contains four columns:

-   y: roi value
-   age: subject's age info
-   sex: subject's sex info
-   icv: subject's intracranial volume

```{r, eval = FALSE}
age_df = data.frame(age_df)
features = colnames(age_df)[c(6:56)]
age = "age"
sex = "sex"
icv = "ICV_baseline"
age_df[[sex]] = as.factor(age_df[[sex]])
```

## Create a list of data sets for all rois

```{r, eval = FALSE}
# Create sub_df for different features
df_var = paste0("sub_df_", 1:length(features))
for (i in 1:length(features)){
  sub_df = age_df[,c(features[i], age, sex, icv)] %>% na.omit() 
  colnames(sub_df) = c("y", "age", "sex", "icv")
  assign(df_var[i], sub_df)
}
```

## Create age trend data sets for all rois

```{r, eval = FALSE}
# For MAC users
age_list = mclapply(1:length(features), function(w){
  age_sub = age_list_gen (sub_df = eval(parse(text = paste0("sub_df_",w))),  lq = 0.25, hq = 0.75)
  return(age_sub)
}, mc.cores = detectCores()) 

# For Windows users
age_list = mclapply(1:length(features), function(w){
  age_sub = age_list_gen (sub_df = eval(parse(text = paste0("sub_df_",w))),  lq = 0.25, hq = 0.75)
  return(age_sub)
}, mc.cores = 1) 

names(age_list) = features

quantile_type = c(paste0("quantile_", 100*0.25), "median", paste0("quantile_", 100*0.75))
ComBatFamQC::age_shiny(age_list, features, quantile_type)
```

# Residual Generation

In this step, we would like to generate different sets of residuals removing batch effect together with specific covariates' effects.

## Get harmonized data set
```{r, eval=FALSE}
features = colnames(adni)[c(8:69)]
covariates = c("timedays", "AGE", "SEX", "DIAGNOSIS")
interaction = c("timedays,DIAGNOSIS")
batch = "manufac"
result_orig = visual_prep(type = "lm", features = features, batch = batch, covariates = covariates, interaction = interaction, smooth = NULL, random = NULL, df = adni)
harmonized_df = result_orig$harmonized_df
```

## Generate residual data set

Specify parameters carefully based on regression type and which covariates' effects to remove.

### Generate residuals from scratch
```{r, eval=FALSE}
result_residual = residual_gen(type = "lm", features = features, batch = batch, covariates = covariates, interaction = interaction, smooth = NULL, df = harmonized_df, rm = c("timedays", "AGE", "SEX", "DIAGNOSIS"))

# save residual data set
write_csv(result_residual$residual, "path/to/save/residual.csv")

# save regression model
writeRDS(result_residual$model, "path/to/save/regression_model.rds")
```

### Generate residuals from existing model
```{r, eval=FALSE}
result_residual = residual_gen(type = "lm", features = features, batch = batch, covariates = covariates, interaction = interaction, smooth = NULL, df = harmonized_df, rm = c("timedays",  "SEX", "DIAGNOSIS"), model = TRUE, model_path = "path/to/save/regression_model.rds")
```